# Analysis of gender distribution of authors of journals articles in musicology

T. Eerola, 29/9/2025, update 30/1/2026 after Dana's manual edits.

This is the first pass to analyse gender distribution of the authors in musicology journals since 2000. The dataset contains 15 journals in musicology that were selected from various [Scimago](https://www.scimagojr.com/journalrank.php?area=1200&category=1210#google_vignette) list minus music psychology journals. The journal selection can be altered, but currently these journals are included (primary sub-discipline in brackets):

- Journal of the American Musicological Society (Musicology)
- Music and Letters (Musicology)                            
- Journal of Musicology (Musicology)                        
- Nineteenth-Century Music Review (Musicology)              
- Nineteenth Century Music (Musicology)                     
- Music Theory Spectrum  (Music Analysis)                        
- Ethnomusicology (Ethnomusicology)                              
- Ethnomusicology Forum (Ethnomusicology)                        
- Music Analysis (Music Analysis)                               
- Popular Music and Society (Popular Music)                   
- Popular Music (Popular Music)                                
- Journal of Popular Music Studies (Popular Music)             
- Journal of World Popular Music (Popular Music)               
- Journal of Popular Music Education (Popular Music)           
- Popular Music History (Popular Music)

This amounted to 4,089 authors.

At the moment, the data is nearly complete although `{r} round(887/4089*100)`% of the affiliations are missing.

```{r}
#| id: libraries
#| warning: false
#| eval: true
#| echo: false
#| message: false
library(tidyverse)
library(stringr)
library(ggplot2)
library(bib2df)
library(countrycode)
library(easystats)
library(here)
```

## Load and preprocess data

```{r}
#| id: preprocess
#| eval: false
#| output: asis
#| warning: false
 
source('scripts/load_scopus_datasets.R') # Creates full_names dataset
source('scripts/attribute_country.R') # Add country affiliations
source('scripts/attribute_gender.R') # Adds gender attributions (precalculated with API)
source('scripts/create_keys.R') # Add keys for authors and study+author
source('scripts/clean_citations_and_OA.R') # Process citations and OA status
source('scripts/attribute_gender_diagnostics.R') # Process citations and OA status
source('scripts/export_gender_data.R')
```



```{r}
#| id: load-data
#| eval: true
#| echo: false
#| output: asis
#| warning: false
#| messages: false

df <- read.csv('data/gender_in_musicology_exported.csv', stringsAsFactors = FALSE,header = TRUE)
df_checked <- read.csv('data/gender_in_musicology_exported - gender_in_musicology_exported_manual_fixes_26_01_2026.csv', stringsAsFactors = FALSE,header = TRUE)
elim <- setdiff(names(df),names(df_checked)) # eliminate columns that are not in the checked data
df <- df %>% dplyr::select(-all_of(elim))
```

### Check manual corrections

```{r}
#| id: check-data
#| eval: true
#| echo: false
#| output: asis
#| warning: false
#| messages: false

cat(paste("Changed first name:",sum(df$first_name != df_checked$first_name,na.rm=T),"\n\n\n"))

cat(paste("Changed last name:",sum(df$last_name != df_checked$last_name,na.rm=T),"\n\n"))

cat(paste("Missing affiliations - original data:", sum(is.na(df$AFFILIATIONS),na.rm=T),"\n\n"))
cat(paste("Missing affiliations - curated data:",sum(is.na(df_checked$AFFILIATIONS),na.rm=T),"\n\n"))

cat(paste("Changed gender attribution:", sum(df$Gender != df_checked$Gender,na.rm=T),"\n\n"))

# Summarise how many missing affiliations there are by journal
S <- df_checked %>%
  group_by(JOURNAL) %>%
  summarise(
    N = n(),
    N_missing = sum(is.na(AFFILIATIONS)),
    pct_missing = N_missing / N * 100
  )
print(knitr::kable(S, digits = 2, caption = 'Missing affiliations by journal'))

# Journal of Musicology, add 6 missing affiliations manually
# Music Analysis, add 30
# Music Theory Spectrum , add 69
# Nineteenth Century Music, add 30

# Use the manually corrected data
df <- df_checked
df$Gender[df$Gender==""]<-"transgender"
df<-dplyr::filter(df,Gender!="transgender") # remove at this stage, make note!
```

## Summarise

```{r}
#| id: summarise
#| eval: true
#| output: asis
#| warning: false
#| messages: false
#| fig.width: 9
#| fig.height: 6
source('scripts/summarise_gender.R') # different summaries
```

## Quantify roles

Using the odds ratio but note that there are relatively few coauthors in the data.

```{r}
#| id: quantify
#| eval: true
#| warning: false
#| message: false
#| output: asis
source('scripts/quantify_authorship.R') # use Odds
source('scripts/citations.R')           # Citations and gender
print(knitr::kable(citestats_all, digits = 2, caption = 'Citations across all authors'))
print(knitr::kable(stats_all))

source('scripts/open_access.R')         # OA and gender
print(knitr::kable(author_OA, digits = 2, caption = 'Open access across all authors'))
```


## Quantify citations and gender

Probably not relevant quantity for these disciplines, but just to see if there are differences.

```{r}
#| id: citations
#| eval: true
#| warning: false
#| output: asis
#| message: false

source('scripts/citations.R')           # Citations and gender
print(knitr::kable(citestats_all, digits = 2, caption = 'Citations across all authors'))
print(knitr::kable(stats_all))

```

## OA and gender?

```{r}
#| id: OA
#| eval: true
#| warning: false
#| message: false
#| output: asis

source('scripts/open_access.R')         # OA and gender
print(knitr::kable(author_OA, digits = 2, caption = 'Open access across all authors'))
```

## Geographical distributions

Do some basic summaries of geographical distributions. Start with countries with most publications overall.

```{r}
#| id: geographical
#| eval: true
#| output: asis

source('scripts/country_table.R') # OK
source('scripts/visualise_gender.R') # OK

```

## Keyword analysis

TBD.



